Multi-scale assessment of a grassland productivity model

### 2020-12-28 manuscript/manuscript.Rmd
some final minor edits
- use spatial scales throughout instead of extent, iterations, etc.
- add caveat on temporal out of sample validation and wouldn't that be nice

A promising method is low dimensional models, which are process models with some simplified components [@choler2010; @choler2011]. For example, a low dimensional model might approximate transpiration to a function of potential evapotranspiration, soil available water, and live vegetation cover along with a single parameter. As opposed to a high dimensional model with multiple functions accounting for leaf area index, stomatal conductance, rooting depth and surface area, etc [@caylor2009; @asbjornsen2011]. The low dimensional model is advantageous since it can generalize across broad regions with relatively few inputs. Yet they are still susceptible to over-fitting to local conditions since parameters or model structure can be tied to specific locations or plant functional groups [@fisher2020]. Thus parameterizing low-dimensional models must be done with care such that they are applicable to a broad area while maintaining an acceptable level of accuracy.

@@

The PhenoGrass model is an ecohydrology model which has interacting state variables for soil water, plant available water, and plant fractional cover [@hufkens2016]. Model inputs are daily precipitation, temperature, potential evapotranspiration (derived from the Hargreaves equation, @hargreaves1985), and solar radiation. The primary output is fractional vegetation cover (fCover). The original model form, derived in Choler et 2010 and Choler et al. 2011, used only temperature and potential evapotranspiration and was parameterized using satellite-derived NDVI data. @hufkens2016 expanded on the original Choler model by incorporating growth and senescence restraints from temperature and solar radiation, and also included a scaling factor to convert PhenoCam G\textsubscript{cc} data to a fractional cover estimate (see supplemental equations). @hufkens2016 evaluated the PhenoGrass model using 14 grassland PhenoCam sites across Western North America with a total of 34 site years. They found the modelled fractional cover correlated well with annual productivity at both a daily and annual timescale. Despite it's name the PhenoGrass model can theoretically apply to any vegetation type with a distinct growth signal in response to precipitation, as hypothesized in the original threshold-delay model [@ogle2004]. Here we use two additional vegetation types, shrubs and agricultural plots, to test how applicable it is beyond grasslands.

@@

We downloaded all Phenocam data with ROIs of the grasslands (GR), shrublands (SH), and agricultural (AG) vegetation types for the years 2012 to 2018, totalling 89 distinct time series  and 463 site-years (Fig. 1, Table S1). As input to the PhenoGrass model we used the 3-day smoothed G\textsubscript{cc} scaled, for each ROI, from 0-1. In the model parameterization each ROI time series is further transformed to a fractional cover estimate using the local mean annual precipitation (MAP) combined with a scaling factor [@hufkens2016; @donohue2013].

@@

To find the most appropriate scale we evaluated the model using three different combinations of ecoregion and vegetation type, with 11 total model parameterizations (Fig. 2). Here we use the term “spatial scale” to refer to the combination of ecoregion/s and vegetation type/s used within each model. This includes using all vegetation types with an ecoregion, or all sites of a specific vegetation type from several ecoregions. The largest scale used all Phenocam locations described above (89 sites). Next were all sites, respectively, within the three vegetation types indicated by the ROI (grasslands, shrublands, and agricultural). Finally, we parameterized models for each vegetation type within each Level 1 North American Ecoregion (eg. All grassland sites within the Great Plains ecoregion). All sets of parameterized models were limited to have at least five sites.

@@

Models which exceed the threshold were subject to further evaluation. For each model we performed a leave one out cross-validation, where the model was re-fit with one Phenocam site not included in the training data, and then evaluated against this left out site. In this step a scaling coefficient to link mean annual precipitation with PhenoCam G\textsubscript{cc} was held constant at the value obtained in first fitting. The resulting \(\overline{NSE}\) and \textit{F} are from all modelled sites using their respective out of sample test.

@@

At the largest scale, where the PhenoGrass model was parameterized using all 89 sites, the model performed poorly with an \(\overline{NSE}\) value among sites of 0.31 (Table 1; Fig. 3). Models built using all sites of a respective vegetation type performed poorly as well, though were slightly better than the all site model (Fig. 3). The best model performance was achieved when models were built using a single vegetation type subset to a single ecoregion. Grasslands within the Great Plains and Eastern Temperate Forests ecoregions were the only instances where \(\overline{NSE}\) exceeded the 0.65 threshold, though Grasslands within N.W. Forests came close (NSE = 0.64).

@@

In all 11 iterations the PhenoGrass model tended to underestimate the highest fCover values, and to a lesser degree overpredict the lowest values (Figs. 3,4). The best performing iterations (Grasslands in the Great Plains and Eastern Temperate forests) minimized this effect (Fig. 4). The worst performing iteration, Grasslands in N.A. Deserts, had little variation in predicted fCover values, resulting in the lowest \(\overline{NSE}\) overall.

The grassland vegetation type, subset to specific ecoregions, predominantly outperformed other iterations of the PhenoGrass model (Table 1). Models built using grasslands within the Eastern Temperate Forest and Great Plains ecoregions had the highest \(\overline{NSE}\) values of 0.82 and  0.69, respectively. Using leave one out cross-validation on these two grassland model iterations resulted in similar errors of 0.79 and 0.67 for the Eastern Temperate Forest and Great Plains, respectively. Though N.W. Forests grasslands had an in-sample \(\overline{NSE}\) just below the 0.65 threshold, the cross-validation was well below it (0.52). Grasslands in the North American deserts were not modelled well at any scale and had the lowest \(\overline{NSE}\) values in the entire analysis. The observed greenness patterns of these desert grasslands had extremely high variability in their magnitude and timing, with short distinct peaks in greenness and numerous off-peak fluctuations. The fitted model, which minimized the mean CVME among the 5 sites, was not able to reproduce this high variability and instead produced fCover values that were severely constrained to a narrow range (Fig. S1).

Agriculture and shrubland sites were poorly modelled at all scales. Performance of agriculture within the E. Temperate Forest ecoregion (NSE = 0.33) improved over the All Agriculture model (NSE = 0.18), but decreased in the Great Plains (from 0.24 to 0.18). There was only a single ecoregion with a minimum of five shrubland sites, N.A. Deserts, and it performed only slightly better than the All Shrubland model. Shrublands in N.A. Deserts did not have the high variability seen in desert grasslands.

@@

We performed an extensive evaluation of the PhenoGrass model across ecoregions and vegetation types to determine the best scale at which to parameterize and apply the model. We found the model most suitable to grassland vegetation when constrained to the ecoregion level, though it did not perform well in grasslands in the North American desert ecoregion. Shrublands and agriculture were not well represented by the model regardless of the scale. Results from this study will facilitate long-term projections of grassland productivity constrained to an appropriate vegetation type and scale.

@@

Shrublands were not well modelled at any scale. Dryland shrubs, representing 8 of the 15 shrubland PhenoCams analysed here, coexist with grasses by accessing different pools of soil water [@weltzin2000; @muldavin2008], thus have different responses to precipitation and resulting greenness patterns [@browning2017; @yan2019]. A prior form of the PhenoGrass model was designed to work with dryland shrubs by using two soil water pools [@ogle2004], yet here PhenoGrass, with a single soil water pool, was less effective for shrubland vegetation. The single pool of the PhenoGrass model is coupled with fluxes from precipitation and evapotranspiration, thus is not well-suited for representing the deeper water pools that shrubs can routinely access [@schenk2002; @ward2013]. Potential improvements would likely need to incorporate a deep soil water pool, in addition to the shallow, which are each utilized by the respective plant functional groups. This has already been implemented in highly parameterized ecohydrology models [@scanlon2005;@lauenroth2014] and could potentially be used here to make a more generalized PhenoShrub model to apply across large scales. This approach could also help in modelling N.A. Desert grasslands which coexist among shrubs.

@@

@hufkens2016 originally evaluated the PhenoGrass model using 14 grassland sites distributed among seven North American ecoregions. In their evaluation they had an average R\textsuperscript{2} of 0.71, while here the model performed poorly when using more than 1 ecoregion. It’s likely that the original 14 grassland sites were ideal locations for the PhenoGrass model, since on average they have a single greenup season every year in the spring or summer (Fig. 5A). The additional 24 grassland sites used in the current study have high seasonal variability and elongated growing seasons (Fig. 5B, Fig. S1), and were thus more difficult to represent in a single continental scale grassland model.

@@

Replication is an important step in the scientific process, especially given newly available data. Here we have validated prior modelling work and highlighted its limitations. Newer small scale vegetation models can be validated in the same framework and applied to areas where PhenoGrass performs poorly. This can result in a spatial ensemble where the output for any one location and vegetation type is represented by the most appropriate model. Our current work will allow for long-term small scale projections of grassland productivity for a large fraction of North America.

>>>>

A promising method for this is low dimensional models, which are process models with some simplified components [@choler2010; @choler2011]. For example, a low dimensional model might approximate transpiration to a function of potential evapotranspiration, soil available water, and live vegetation cover along with a single parameter. As opposed to a high dimensional model with multiple functions accounting for leaf area index, stomatal conductance, rooting depth and surface area, etc [@caylor2009; @asbjornsen2011]. The low dimensional model is advantageous since it can generalize across broad regions with relatively few inputs. Yet they are still susceptible to over-fitting to local conditions since parameters or model structure can be tied to specific locations or plant functional groups [@fisher2020]. Thus parameterizing low-dimensional models must be done with care such that they are applicable to a broad area while maintaining an acceptable level of accuracy.

@@

The PhenoGrass model is an ecohydrology model which has interacting state variables for soil water, plant available water, and plant fractional cover [@hufkens2016]. Model inputs are daily precipitation, temperature, potential evapotranspiration (derived from the Hargreaves equation, @hargreaves1985), and solar radiation. The primary output is fractional vegetation cover (fCover). The original model form, derived in Choler et 2010 and Choler et al. 2011, used only temperature and potential evapotranspiration and was parameterized using satellite-derived NDVI data. @hufkens2016 expanded on the original Choler model by incorporating growth and senescence restraints from temperature and solar radiation, and also included a scaling factor to convert PhenoCam G\textsubscript{cc} data to a fractional cover estimate (see supplemental equations). @hufkens2016 evaluated the PhenoGrass model using 14 grassland PhenoCam sites across Western North America with a total of 34 site years. They found the modelled fractional cover correlated well with annual productivity at both a daily and annual timescale. Despite it's name the PhenoGrass model can theoretically apply to any vegetation type with a distinct growth signal in response to precipitation, as hypothesized in the original threshold-delay model [@ogle2004]. Here we evaluate two additional vegetation types, shrubs and agricultural plots, to test how applicable it is beyond grasslands.

@@

We downloaded all Phenocam data with ROIs of the grasslands (GR), shrublands (SH), and agricultural (AG) vegetation types for the years 2012 to 2018, totalling 89 distinct time series and 463 site-years (Fig. 1, Table S2). As input to the PhenoGrass model we used the 3-day smoothed G\textsubscript{cc} scaled, for each ROI, from 0-1. In the model parameterization each ROI time series is further transformed to fractional vegetation cover (\textit{fCover}, see supplmental equations) using the local mean annual precipitation (MAP) combined with a scaling factor [@hufkens2016; @donohue2013].

@@

To find the most appropriate spatial scale we evaluated the model using three different combinations of ecoregion and vegetation type, with 11 total model parameterizations (Fig. 2). Here we use the term “spatial scale” to refer to the combination of ecoregion/s and vegetation type/s used within each model. This includes using all sites of one vegetation types within an ecoregion, or all sites of a specific vegetation type from several ecoregions (Fig. 2). The largest spatial scale used all Phenocam locations described above (89 sites). Next were all sites, respectively, within the three vegetation types indicated by the ROI (grasslands, shrublands, and agricultural). Finally, we parameterized models for each vegetation type within each Level 1 North American Ecoregion (eg. All grassland sites within the Great Plains ecoregion). All sets of parameterized models were limited to have at least five sites.

@@

Models which exceeded the threshold were subject to further evaluation. For each model we performed a leave one out cross-validation, where the model was re-fit with one Phenocam site not included in the training data, and then evaluated against this left out site. In this step a scaling coefficient to link mean annual precipitation with PhenoCam G\textsubscript{cc} was held constant at the value obtained in first fitting. The resulting \(\overline{NSE}\) and \textit{F} are from all modelled sites using their respective out of sample test.

@@

At the largest spatial scale, where the PhenoGrass model was parameterized using all 89 sites, the model performed poorly with an \(\overline{NSE}\) value among sites of 0.31 (Table 1; Fig. 3). Models built using all sites of a respective vegetation type performed poorly as well, though were slightly better than the all site model (Fig. 3). The best model performance was achieved when models were built using a single vegetation type subset to a single ecoregion. Grasslands within the Great Plains and Eastern Temperate Forests ecoregions were the only instances where \(\overline{NSE}\) exceeded the 0.65 threshold, though Grasslands within N.W. Forests came close (NSE = 0.64).

@@

In all 11 models the PhenoGrass model tended to underestimate the highest fCover values, and to a lesser degree overpredict the lowest values (Figs. 3,4). The best performing models (Grasslands in the Great Plains and Eastern Temperate forests) minimized this effect (Fig. 4). The worst performing models, Grasslands in N.A. Deserts, had little variation in predicted fCover values, resulting in the lowest \(\overline{NSE}\) overall.

The grassland vegetation type, subset to specific ecoregions, predominantly outperformed models built with other spatial scales (Table 1). Models built using grasslands within the Eastern Temperate Forest and Great Plains ecoregions had the highest \(\overline{NSE}\) values of 0.82 and 0.69, respectively. Using leave one out cross-validation on these two grassland models resulted in similar errors of 0.79 and 0.67 for the Eastern Temperate Forest and Great Plains, respectively. Though N.W. Forests grasslands had an in-sample \(\overline{NSE}\) just below the 0.65 threshold, the cross-validation was well below it (0.52). Grasslands in the North American deserts were not modelled well at any spatial scale and had the lowest \(\overline{NSE}\) values in the entire analysis. The observed greenness patterns of these desert grasslands had extremely high variability in their magnitude and timing, with short distinct peaks in greenness and numerous off-peak fluctuations. The fitted model, which minimized the mean CVME among the 5 sites, was not able to reproduce this high variability and instead produced fCover values that were severely constrained to a narrow range (Fig. S1).

Agriculture and shrubland sites were poorly modelled at all spatial scales. Performance of agriculture within the E. Temperate Forest ecoregion (\(\overline{NSE}\) = 0.33) improved over the All Agriculture model (\(\overline{NSE}\) = 0.18), but decreased in the Great Plains (\(\overline{NSE}\) from 0.24 to 0.18). There was only a single ecoregion with a minimum of five shrubland sites, N.A. Deserts, and it performed only slightly better than the All Shrubland model. Shrublands in N.A. Deserts did not have the high variability seen in desert grasslands.

@@

We performed an extensive evaluation of the PhenoGrass model across ecoregions and vegetation types to determine the best spatial scale at which to parameterize and apply the model. We found the model most suitable to grassland vegetation when constrained to the ecoregion level, though it did not perform well in grasslands in the North American desert ecoregion. Shrublands and agriculture were not well represented by the model regardless of the spatial scale. Results from this study will facilitate long-term projections of grassland productivity constrained to an appropriate vegetation type and scale.

@@

Shrublands were not well modelled at any spatial scale. Dryland shrubs, representing 8 of the 15 shrubland PhenoCams analysed here, coexist with grasses by accessing different pools of soil water [@weltzin2000; @muldavin2008], thus have different responses to precipitation and resulting greenness patterns [@browning2017; @yan2019]. A prior form of the PhenoGrass model was designed to work with dryland shrubs by using two soil water pools [@ogle2004], yet here PhenoGrass, with a single soil water pool, was less effective for shrubland vegetation. The single pool of the PhenoGrass model is coupled with fluxes from precipitation and evapotranspiration, thus is not well-suited for representing the deeper water pools that shrubs can routinely access [@schenk2002; @ward2013]. Potential improvements would likely need to incorporate a deep soil water pool, in addition to the shallow, which are each utilized by the respective plant functional groups. This has already been implemented in highly parameterized ecohydrology models [@scanlon2005;@lauenroth2014] and could potentially be used here to make a more generalized PhenoShrub model to apply across large scales. This approach could also help in modelling N.A. Desert grasslands which coexist among shrubs.

@@

@hufkens2016 originally evaluated the PhenoGrass model using 14 grassland sites distributed among seven North American ecoregions. In their evaluation they had an average R\textsuperscript{2} of 0.71, while here the model performed poorly when using more than 1 ecoregion. It’s likely that the original 14 grassland sites were ideal locations for the PhenoGrass model, since on average they have a single greenup season every year in the spring or summer (Fig. 5A). The additional 24 grassland sites used in the current study have high seasonal variability and elongated growing seasons (Fig. 5B, Fig. S1), and were thus more difficult to represent in a single continental scale grassland model. This highlights the need for longer time series in evaluating low-dimensional models as it may take many years for a single location to experience the full range of variability. As the PhenoCam data archive grows, temporally out of sample validation can be done to better evaluate performance into novel conditions.

@@

Replication is an important step in the scientific process, especially given newly available data. Here we have validated prior modelling work and highlighted its limitations. Newer vegetation models can be validated in the same framework and applied to areas where PhenoGrass performs poorly. This can result in a spatial ensemble where the output for any one location and vegetation type is represented by the most appropriate model. Our current work will allow for long-term projections of grassland productivity for a large fraction of North America.
